### 2016-06-13 Projects/DelCo_2014_2015/Reports/Hose_Stream_Tests/TeX_Files/DelCo_Hose_Stream_Tests.tex
minor updates to text

One of the objectives of the Fire Research Division at the National Institute of Standards and Technology (NIST) is to improve the safety and effectiveness of firefighters through improved knowledge of fire behavior and examine how different firefighting tactics affect the fire environment. NIST and UL FSRI have conducted a significant amount of research examining how ventilation affects the growth and spread of fire within structures and how the air flow to the fire may be controlled to limit or delay the growth of the fire~\cite{madrzykowski2009fire,kerber2009fire}. The studies have resulted in guidance to the fire service regarding ventilation tactics. However, ventilation tactics alone will not result in the extinguishment of the fire; fire suppression with hose streams is also needed.

Fire suppression tactics using hose streams also affect the ventilation in a structure and may impact the movement of smoke and heat through a structure. As vents are made to advance the hoseline or if ventilation inducing handline tactics are in practice, hose stream impacts on ventilation may be even more significant. Additional research addressing the coordination of suppression tactics and the impact on ventilation is needed to complete recommendations on fire control tactics to appropriate standards, education, and training documents.

There exists a multitude of different tools, methods, and techniques used to apply water for fire suppression using hose streams. A common type of nozzle that is used to apply water during fire suppression in residential structures is an adjustable-pattern spray nozzle, commonly referred to as a combination nozzle~\cite{NFPA_1964}. Most combination nozzles contain an adjustable tip that can be rotated to produce different stream patterns. The types of streams produced by a combination nozzle are often divided into three categories based on the angle of the stream with respect to the nozzle: a straight stream typically produces an angle less than 15\SI{}{\degree}; a narrow fog stream produces an angle between 15\SI{}{\degree} and 45\SI{}{\degree}; and a wide fog stream produces an angle between 45\SI{}{\degree} and 80\SI{}{\degree}~\cite{IFSTA:Essentials_of_FF}. Increasing the angle of the stream allows for a larger area to be covered by the water spray.  However, increasing the angle decreases the distance reached by the stream and its penetration power. 
%\textit{Essentials of Fire Fighting} defines 
%https://docs.google.com/presentation/d/1hZ5DZhYXxCkcIEHNZKumH48b_Q53iM2D7asXkMYlVMM/embed?hl=en&size=l&slide=id.p39

%http://www.tft.com/literature/library/files/bfa.pdf
%Keith Royer, "Water for Fire Fighting", (Engineering Extension Service, Iowa State University, Ames, Iowa) p. 6 
In addition to the type of stream pattern used, the method used to apply the hose stream can also vary. For example, Royer suggested using a clockwise rotation instead of a counterclockwise application pattern for the following reasons~\cite{Royer:ISU}:
\begin{enumerate} 
	\item Clockwise rotation drives most of the heated gases, smoke, and flames away from the nozzle, while counterclockwise rotation does just the opposite.
	\item Steam formed by the clockwise rotation has a violent rolling action. Counterclockwise rotation produces steam with an inactive and lazy action.
	\item A clockwise rotation increases the efficiency of water and produces a faster knockdown time than counterclockwise rotation.
\end{enumerate}

\section{Objectives}
\label{sec:objectives}
This report is concerned with examining the impact of hose stream and application pattern, or nozzle motion, selection on air movement inside two full scale test structures built to represent residential sized structures. Numerous experiments containing different flow path configurations were conducted to study the impact of hose stream and water application selection on ventilation in a structure. More specifically, the experiments were conducted to fulfill the following objectives:
\begin{enumerate}
	\item Determine which hose stream patterns and application patterns cause significant air movement in a structure 
	\item Determine the amount of air movement induced by different hose streams and application patterns

\end{enumerate}
Four types of hose streams and four different water application patterns were studied throughout the experiments described in this report. Three types of hose stream patterns using a combination nozzle were studied: straight stream, narrow fog stream, and wide fog stream. Water was flowed into the structure by having the hose in a static, or fixed, position; by moving the hoseline left-to-right across the room in a sweeping motion; and by rotating the hoseline in both the clockwise and counterclockwise directions. Additionally, a set of experiments comparing the impact of a straight stream from a combination nozzle and a solid stream from a smooth bore nozzle equipped with a 1~in tip on air flows in a structure was conducted. Images of the four types of hose streams are presented in Fig.~\ref{fig:hose_streams}.

@@

\section{Experimental Setup}
\label{sec:exp_setup}
The series of field experiments described in this report were conducted in two structures of similar design located at the Delaware County Emergency Services Training Center (ESTC) in Sharon Hill, PA. Differential pressure and temperature sensors were installed at ventilation points throughout the residential sized test structures to determine air velocity during the water flow experiments.

@@

\subsubsection{Construction}
\label{sec:construction}
Each test structure was built on a concrete slab as shown in Fig.~\ref{fig:struct_pics}. The Single Story Structure was designed to simulate a single-story residential structure, and the Two Story Structure was designed to simulate a two-story residential structure. The first floor of each structure had an outer wall composed of interlocking concrete blocks with equal side lengths of 0.61~m (2~ft). The joints and gaps between the blocks were filled with high temperature insulation.

@@

The exterior doorways of each structure and the stairwell doorway on the second level of the Two Story Structure all contained doors that were opened or closed at certain instances during tests to change the ventilation pattern of the structure. All other doorways in the structures did not contain a door. In order to close these doorways, a sheet of gypsum board was used to cover the opening, and the doorway remained closed for the duration of the test procedure.
\clearpage

@@

\subsection{Instrumentation}
\label{sec:instrumentation}
For every experiment, gas velocity at an interior doorway was calculated using differential pressure transducers connected to bi-directional velocity probes~\cite{McCaffrey:Combustion_and_Flame}. An array of six bi-directional probes located at the northeast interior doorway of the Single Story Structure is pictured below in Fig.~\ref{fig:BDPs}. A single thermocouple was attached to each bi-directional probes to measure the gas temperature to be used in conjunction with the pressure measurement to derive velocity. Every thermocouple attached to the bi-directional probes was an exposed-bead, Chromel-Alumel (type K) with a 1.0~mm (0.04 in) diameter. The thermocouple wire was sheathed in a 3.2~mm (0.13 in) diameter Inconel shield that started at the exposed bead and was 0.76~m (2.5~ft) in length.

@@

\subsubsection{Length Measurements}
Each length measurement was taken carefully. Length measurements such as the room dimensions, instrumentation array locations, and fire apparatus (e.g., nozzle, sprinkler, or fan) placement were made with a hand held laser measurement device which has an accuracy of $\pm$6.0~mm (0.25~in) over a range of 0.61~m (2.0~ft) to 15.3~m (50.0~ft)~\cite{StanleyTools}.  However, conditions affecting the measurement, such as levelness of the device, yields an estimated uncertainty of $\pm$0.5\% for measurements in the 2.0~m (6.6~ft) to 10.0~m (32.8~ft) range. Steel measuring tapes with a resolution of $\pm$0.5~mm (0.02~in) were used to locate individual sensors within a measurement array and to measure and position the furniture. The steel measuring tapes were manufactured in compliance with NIST Manual 44, which specifies a tolerance of $\pm$1.6~mm (0.06~in) for 9.1~m (30~ft) tapes and $\pm$6.4~mm (0.25~in) for 30.5~m (100~ft) tapes~\cite{Butcher:2012}. Some issues, such as ``soft'' edges on the upholstered furniture, result in an estimated total expanded uncertainty of $\pm$1.0\%.

\subsubsection{Bi-Directional Probes}
Bi-directional probes with pressure transducers and single thermocouples were used to measure the gas velocity. Bare-bead, type K thermocouples were co-located with the probe. A gas velocity measurement study examining the doorway flow of pre-flashover compartment fires yielded expanded uncertainty measurements ranging from $\pm0.14$ to $\pm0.22$ for bi-directional probes of similar design~\cite{Bryant:FSJ2009}. The total expanded uncertainty for gas velocity in these experiments is estimated to be $\pm18$\%.

>>>>

One of the objectives of the Fire Research Division at the National Institute of Standards and Technology (NIST) is to improve the safety and effectiveness of firefighters through improved knowledge of fire behavior and examination of how different firefighting tactics affect the fire environment. NIST has conducted a significant amount of research examining how ventilation affects the growth and spread of fire within structures and how the air flow to the fire may be controlled to limit or delay the growth of the fire~\cite{madrzykowski2009fire,kerber2009fire}. The studies have resulted in guidance to the fire service regarding ventilation tactics. However, ventilation tactics alone will not extinguish the fire; fire suppression with hose streams is also needed.

Fire suppression tactics using hose streams also affect the ventilation in a structure and may impact the movement of smoke and heat through a structure. If vents are made to advance the hoseline or if ventilation inducing handline tactics are in practice, hose stream impacts on ventilation may be even more significant. Additional research addressing the coordination of suppression tactics and the impact on ventilation is needed to complete recommendations on fire control tactics to appropriate standards, education, and training documents.

There exists a multitude of different tools, methods, and techniques used to apply water for fire suppression using hose streams. A common type of nozzle that is used to apply water during fire suppression in residential structures is an adjustable-pattern spray nozzle, commonly referred to as a combination nozzle~\cite{NFPA_1964}. Most combination nozzles contain an adjustable tip that can be rotated to produce different stream patterns. The types of streams produced by a combination nozzle are often divided into three categories based on the angle of the stream with respect to the nozzle: a straight stream typically produces an angle less than 15\SI{}{\degree}; a narrow fog stream produces an angle between 15\SI{}{\degree} and 45\SI{}{\degree}; and a wide fog stream produces an angle between 45\SI{}{\degree} and 80\SI{}{\degree}~\cite{IFSTA:Essentials_of_FF}. Increasing the angle of the stream allows for a larger area to be covered by the water spray.  However, increasing the angle decreases the distance reached by the stream and its penetration power.

In addition to the type of stream pattern used, the method used to apply the hose stream can also vary. For example, Royer suggested using a clockwise rotation instead of a counterclockwise application pattern for the following reasons~\cite{Royer:ISU}:
\begin{enumerate} 
	\item Clockwise rotation drives most of the heated gases, smoke, and flames away from the nozzle, while counterclockwise rotation does just the opposite.
	\item Steam formed by the clockwise rotation has a violent rolling action. Counterclockwise rotation produces steam with an inactive and lazy action.
	\item A clockwise rotation increases the efficiency of water and produces a faster knockdown time than counterclockwise rotation.
\end{enumerate}

\section{Objectives}
\label{sec:objectives}
This report focuses on examining the impact of hose stream and application pattern (or nozzle motion) selection on air movement inside two full-scale test structures built to represent residential sized structures. Experiments containing different flow path configurations were conducted to study the impact of hose stream and water application selection on ventilation in a structure. More specifically, the experiments were conducted to fulfill the following objectives:
\begin{enumerate}
	\item Determine which hose stream patterns and application patterns cause significant air movement in a structure 
	\item Determine the amount of air movement induced by different hose streams and application patterns

\end{enumerate}
Four types of hose streams and four water application patterns were studied as part of the experiments described in this report. Three of the hose stream patterns studied were from a combination nozzle: straight stream, narrow fog stream, and wide fog stream. Water was flowed into the structure by having the hose in a static, or fixed, position; by moving the hoseline left-to-right across the room in a sweeping motion; and by rotating the hoseline in both the clockwise and counterclockwise directions. Additionally, experiments comparing the differences in impact on air flows in a structure from a combination nozzle straight stream and a solid stream from a smooth bore nozzle equipped with a 1~in tip were conducted. Images of the four types of hose streams are presented in Fig.~\ref{fig:hose_streams}.

@@

\section{Experimental Setup}
\label{sec:exp_setup}
The series of field experiments described in this report were conducted in two structures of similar design located at the Delaware County Emergency Services Training Center (ESTC) in Sharon Hill, PA. Differential pressure and temperature sensors were installed at ventilation points throughout the test structures to determine air velocity during the water flow experiments.

@@

\subsubsection{Construction}
\label{sec:construction}
Each test structure was built on a concrete slab as shown in Fig.~\ref{fig:struct_pics}. The Single Story Structure was designed to simulate a single-story residential structure, and the Two Story Structure was designed to simulate a two-story residential structure. The first floor of each structure had an outer wall composed of interlocking concrete blocks of equal side lengths of 0.61~m (2~ft). The joints and gaps between the blocks were filled with high temperature insulation.

@@

The exterior doorways of each structure and the stairwell doorway on the second level of the Two Story Structure all contained doors that were opened or closed at certain instances during tests to change the ventilation pathways within the structure. All other doorways in the structures did not contain a door. In order to close these doorways, a sheet of gypsum board was used to cover the opening, and the doorway remained closed for the duration of the test procedure.
\clearpage

@@

\subsection{Instrumentation}
\label{sec:instrumentation}
For every experiment, gas velocity at an interior doorway was calculated using differential pressure transducers connected to bi-directional velocity probes~\cite{McCaffrey:Combustion_and_Flame}. An array of six bi-directional probes located at the northeast interior doorway of the Single Story Structure is pictured below in Fig.~\ref{fig:BDPs}. A single thermocouple was attached to each bi-directional probes to measure the gas temperature to be used in conjunction with the pressure measurement to calculate velocity. Every thermocouple attached to the bi-directional probes was an exposed-bead, Chromel-Alumel (type K) with a 1.0~mm (0.04 in) diameter. The thermocouple wire was sheathed in a 3.2~mm (0.13 in) diameter Inconel shield that started at the exposed bead and was 0.76~m (2.5~ft) in length.

@@

\subsubsection{Length Measurements}
Each length measurement was taken carefully. Length measurements such as the room dimensions, instrumentation array locations, and fire apparatus (e.g., nozzle, sprinkler, or fan) placement were made with a hand held laser measurement device which has an accuracy of $\pm$6.0~mm (0.25~in) over a range of 0.61~m (2.0~ft) to 15.3~m (50.0~ft)~\cite{StanleyTools}.  However, conditions affecting the measurement, such as levelness of the device, yields an estimated uncertainty of $\pm$0.5\% for measurements in the 2.0~m (6.6~ft) to 10.0~m (32.8~ft) range. Steel measuring tapes with a resolution of $\pm$0.5~mm (0.02~in) were used to locate individual sensors within a measurement array. The steel measuring tapes were manufactured in compliance with NIST Manual 44, which specifies a tolerance of $\pm$1.6~mm (0.06~in) for 9.1~m (30~ft) tapes and $\pm$6.4~mm (0.25~in) for 30.5~m (100~ft) tapes~\cite{Butcher:2012}.

\subsubsection{Bi-Directional Probes}
Bi-directional probes with pressure transducers and single thermocouples were used to measure the gas velocity. Inconel-sheathed, exposed bead, type K thermocouples were co-located with each probe. A gas velocity measurement study examining the doorway flow of pre-flashover compartment fires yielded expanded uncertainty measurements ranging from $\pm0.14$ to $\pm0.22$ for bi-directional probes of similar design~\cite{Bryant:FSJ2009}. The total expanded uncertainty for gas velocity in these experiments is estimated to be $\pm18$\%.
